Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-15 paper.tex
Update

\jkcomment{If we're going to have this paragraph of ARG bashing then
I guess we'll have to do more than just single out Iain and Aylwyn.}
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

@@

\begin{figure}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

\subsection*{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

@@

We support changing the simulation model, so that we can easily
run hybrid simulations like described by~\cite{bhaskar2014distortion}
This ia a powerful approach because it lets us combine any
number of different simulation models.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/mutation_perf}
\caption{\label{fig-mutation-perf} Time required to run
\texttt{sim\_mutations} on tree sequences generated
by \texttt{sim\_ancestry} (with a population size of $10^4$
and recombination rate of $10^{-8}$) for varying (haploid) sample
size and sequence length. We run 10 replicate mutation simulations
each for three different mutation rates, and report the average
CPU time required (Intel Core i7-9700).
(A) Holding sequence length fixed at 10 megabases and varying the
number of samples (tree tips) from 10 to 100,000.
(B) Holding number of samples fixed at 1000, and varying the sequence
length from 1 to 100 megabases.}
\end{figure}

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
opens the possibility for new applications of coalescent simulations.
Also \stdpopsim\ is cool~\citep{adrion2020community}

>>>>

However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
Others state that an ARG is equivalent to the sequence of marginal
% Also SARGE paper
trees along the genome~\citep{mirzaei2016rent}

Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

\subsection*{Discrete time Wright-Fisher}

The assumptions of the classical coalescent model are that both sample size and
the simulated sequence length are small. As modern datasets often include
whole-genome sequences for tens or hundreds of thousands of samples, simulating
such data requires breaking these assumptions. While the coalescent remains a
good approximation to the Wright-Fisher model in the distant past, it can
exhibit large biases in the recent history. Identity-by-descent (IBD),
long-range linkage disequilibrium (LD), and ancestry patterns deviate from
Wright-Fisher expectations, and this bias grows considerably when simulating
larger sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.

A backwards-in-time Wright-Fisher model is supported in \msprime, which is
shown to properly capture the expected distributions of IBD and long-range LD,
as well as ancestry patterns in admixed
populations~\citep{nelson2020accounting}. In short, the classical coalescent
draws recombination events from exponential waiting times, which are continuous
in time, and each recombination then splits a segment into two distinct
ancestors. For large sequence lengths, this can dramatically overestimate the
number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
model iterates through discrete generations so that each gamete has a unique
diploid parent, and multiple recombinations within a generation results in
crossover events between the same two parental haploid copies.  This method is
described in more detail in~\cite{nelson2020accounting}.

@@

We support changing the simulation model at specified times so that we can
easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
can combine any number of different simulation models, allowing for the
flexible choice of simulation scenarios without severely damaging efficiency.
As the discrete time Wright-Fisher model improves accuracy of genealogical
patterns in the recent past, we can simulate the recent history using this
model and then switch to the standard coalescent to more efficiently simulate
the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
hybrid approach provides large efficiency gains.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/mutation-perf}
\caption{\label{fig-mutation-perf} Time required to run
\texttt{sim\_mutations} on tree sequences generated
by \texttt{sim\_ancestry} (with a population size of $10^4$
and recombination rate of $10^{-8}$) for varying (haploid) sample
size and sequence length. We run 10 replicate mutation simulations
each for three different mutation rates, and report the average
CPU time required (Intel Core i7-9700).
(A) Holding sequence length fixed at 10 megabases and varying the
number of samples (tree tips) from 10 to 100,000.
(B) Holding number of samples fixed at 1000, and varying the sequence
length from 1 to 100 megabases.}
\end{figure}

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
% also SARGE
opens the possibility for new applications of coalescent simulations.
With lots of applications, e.g., [cite Osmond and Coop].
Also \stdpopsim\ is cool~\citep{adrion2020community}
